# Reproducible Research
## Peer-graded Assignment: Course Project 1

---
title: "Reproducible Research: Peer Assessment 1"
author: "Valentina Grouverman"
date: "4/15/2018"
output: html_document
---
### Project description

Project dataset: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

The dataset is stored in a comma-separated-value (CSV) file and 
there are a total of 17,568 observations in this dataset.

The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. date: The date on which the measurement was taken in YYY-MM-DD format
3. interval: Identifier for the 5-minute interval in which measument was taken

### Downloading and reading data

```{r loadingReadingData, echo=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destZip = "./activity.zip"

if(!file.exists("./activity.csv")){
     download.file(fileUrl, destfile = destZip) 
     unzip(destZip)
     file.remove(destZip)
}

if(!exists("Activity")){
     Activity <- read.csv("./activity.csv", header = TRUE)
}
```

### Q1: What is mean total number of steps taken per day?
#### Calculate and report the mean and median of the total number of steps taken per day.
```{r Q1, echo=TRUE}
ActivitySplitDate <- split(Activity, as.factor(Activity$date))
Steps_Date <- sapply(ActivitySplitDate, function(x) sum(x$steps, na.rm=T))

mean_display = round(mean(sapply(ActivitySplitDate, function(x) sum(x$steps, na.rm = TRUE))))
median_display = round(median(sapply(ActivitySplitDate, function(x) sum(x$steps, na.rm = TRUE))))

hist(Steps_Date, 
     main = "Number of steps by Date",
     xlab = "Number of steps", 
     col = "grey",
     breaks = 30,
     cex.main = 1)

abline(v = mean_display, lwd = 2, col = "blue")
abline(v = median_display, lwd = 2, col = "green")

legend("topright", 
       lty = 1,
       lwd = 2,
       col = c("blue","green"),
       cex = .5,
       legend = c(paste("Mean equals ", mean_display),
                  paste("Median equals ", median_display))
)
```

### Q2: What is the average daily activity pattern?
#### Which 5-minute interval, on averageacross all the days in the dataset, contains the maximum number of steps?
```{r Q2, echo=TRUE}
library(data.table)
ActivityTbl = data.table(Activity)
Steps_Interval = 
     ActivityTbl[, list(StepsMean = mean(steps, na.rm = T)), by = interval]

with(Steps_Interval,
     plot(interval, 
          StepsMean, 
          type = "l",
          lwd = 1.5,
          main = "Average number of steps by Interval",
          xlab = "5-minute Interval",
          ylab = "Average number of steps across all days")
)

## to get a recod with max value of the StepsMean
MaxStepsInterval = Steps_Interval[which.max(StepsMean), ]
## to display it on the plot
points(MaxStepsInterval$interval, 
       MaxStepsInterval$StepsMean, 
       col="red",
       pch = 19)

legend("topright",
       legend = 
            c(paste("5-minute interval ",MaxStepsInterval$interval), 
              paste("with maximum number of steps ",round(MaxStepsInterval$StepsMean))),
       cex = .5
)
```

### Q3: Imputing missing values
#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r Q3, echo=TRUE}
## to merge tables by interval
setkey(ActivityTbl, interval)
setkey(Steps_Interval, interval)

## to incorporate StepsMean value to the table
MisStepsTbl = ActivityTbl[Steps_Interval]
MisStepsTbl$steps <- ifelse(test = is.na(MisStepsTbl$steps), 
                            yes = round(MisStepsTbl$StepsMean), 
                            no = MisStepsTbl$steps)

## to confirm that we don't have NA in steps
# sum(is.na(MisStepsTbl$steps))

AlteredStepsSplitDate <- split(MisStepsTbl, as.factor(MisStepsTbl$date))
AlteredSteps_Date <- sapply(AlteredStepsSplitDate, function(x) sum(x$steps, na.rm=T))

mean_display = round(mean(sapply(AlteredStepsSplitDate, function(x) sum(x$steps, na.rm = TRUE))))
median_display = round(median(sapply(AlteredStepsSplitDate, function(x) sum(x$steps, na.rm = TRUE))))

hist(AlteredSteps_Date, 
     main = "Imputing missing values: number of steps by Date",
     xlab = "Number of steps", 
     col = "grey",
     breaks = 30,
     cex.main = 1)

abline(v = mean_display, lwd = 2, col = "blue")
abline(v = median_display, lwd = 2, col = "green")

legend("topright", 
       lty = 1,
       lwd = 2,
       col = c("blue","green"),
       cex = .5,
       legend = c(paste("Mean equals ", mean_display),
                  paste("Median equals ", median_display))
)
```

##### Conclusion: 
After replacing missing values with the mean value of the relevant interval, the mean and the median became almost equal to each other.

### Q4: Are there differences in activity patterns between weekdays and weekends?
```{r Q4, echo=TRUE}
MisStepsTbl$week_days = weekdays(as.Date(MisStepsTbl$date))
MisStepsTbl$isWeekend <- ifelse(test = (MisStepsTbl$week_days %in% c("Saturday","Sunday")), 
                                   yes = "Weekends", 
                                   no = "Weekdays")
Steps_IntervalDate = 
     MisStepsTbl[, list(StepsMean = mean(steps, na.rm = T)), by = list(interval,isWeekend)]
     
library(lattice)
xyplot(StepsMean ~ interval|isWeekend,
       data = MisStepsTbl,
       type = "l",
       xlab = "Intervals",
       ylab = "Average number of steps",
       layout = c(1,2))

xyplot(StepsMean ~ interval|isWeekend,
       data = MisStepsTbl,
       type = "l",
       xlab = "Intervals",
       ylab = "Average number of steps",
       layout = c(2,1))

```

##### Conclusion: 
Plots don't show a lot of differences.
